```{r echo=FALSE}
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)
library(ggplot2)
```
###3. 2013-2015 Q1 United States Import/Export Data

The last dataset I chose was US trade data, contributed by Jason Joseph. Jason
was kind enough to provide a .csv file, so I'll load that in directly.

```{r}
trade_data <- read.csv("International Trade in Goods and Services.csv", 
                          stringsAsFactors = FALSE)
tbl_df(trade_data)
```

In spite of the fact that this is the smallest data set I am working with on this
project, it is also probably the most challenging to tidy. There are blank rows,
two header rows, and double row records.

The easiest issue to deal with are the blank rows, so I'll delete them out first
by subsetting out any row not blank in the X2013 column.

```{r}
trade_data <- subset(trade_data, X2013 != "")
tbl_df(trade_data)
```

Now that the empty space has been eliminated, I'll turn my attention to fixing
the fact that the observations span two columns. I can do this by filling in
the same month as the one above it.

```{r}
trade_data$Period[c(3,5,7,9)] <- c("January", "February","March","April")
tbl_df(trade_data)
```

These column names aren't very descriptive, but I can tell what they should be
using context clues, so I will fix those now. This also allows me to fix the
issue that there are two header rows. It creates another aspect of "untidyness"
by combining two variables into one column, but I'm not really chaning the data,
I'm just relabelling at this point.

```{r}
colnames(trade_data)[2] <- "goods_services"
colnames(trade_data)[3:8] <- c("2013 Exports","2013 Imports","2014 Exports",
                               "2014 Imports","2015 Exports", "2015 Imports")
trade_data <- trade_data[2:9,]
rownames(trade_data) <- 1:8
tbl_df(trade_data)
```


I want to do three things here. First, I want to narrow and lengthen the data
by eliminating the Year/Type columns and putting them in a variable field.
Since this also constitutes an example of more than one variable in a field,
I'll use separate to split them out using regex. Finally, the goods and services
are swung out as column names.

```{r}
tidy_trade <- trade_data %>%
                      gather("Year/Type", "Total", -Period, -goods_services)%>%  
                      separate("Year/Type", 
                                into = c("year", "import_export"), 
                                sep = "[[:space:]]")%>%
                      spread(goods_services, Total)
tbl_df(tidy_trade)
```

It looks good, but I kind of went back and forth as to whether Goods and Services
should be spread out as separate columns the way I did here. After all, it is just
a two level categorical value, just like import/export. Should the same be done 
with that variable?

I decided no because it would require combining import/exports with goods and
services. "Goods Imported" would be a column, for instance. While this would 
be a nice tabular display of the data, it seems like it would violate the variable
combination rule, so I abstained.

However, there is a bit more work to be done. Since the csv contained commas, the values
were imported as strings, therefore I am unable to compute at this point. The 
commas also prevent me from coercing directly to numeric (I thought R might be
able to figure this out, but I was wrong). In any event, it's a good opportunity to
practice Regex a bit more.

```{r}
tidy_trade$'Goods ' <- as.numeric(str_replace(tidy_trade$'Goods ', ",", ""))
tidy_trade$Services  <- as.numeric(str_replace(tidy_trade$Services , ",", ""))
```

Before I do the computations Jason inquired about, I'm going to split up imports
and exports into two separate data frames to make the subsetting a little easier.

```{r}
exports <- subset(tidy_trade, import_export == 'Exports')
imports <- subset(tidy_trade, import_export == 'Imports')
```

With that in hand, I can quickly and easily find the average accross months.

```{r}
#1. average goods exported for each month
mean(exports$Goods)

#2. avegage goods imported for each month
mean(imports$Goods)

#3. average services exported for each month
mean(exports$Services)

#4. average services imported for each month
mean(imports$Services)
```