---
title: "Week 10 Assignment"
author: "Chris Fenton"
date: "November 1, 2015"
output: html_document
---

I'm going to work with the jsonlite package, as it combines sending the HTTP request
and converting to json in one step. 

```{r}
library(jsonlite)
```

I'm going to be working with the NY Times Movie Reviews API. The documentation
can be found [here](http://developer.nytimes.com/docs/movie_reviews_api/).

Just to provide some sort of guiding question, I am going to be comparing Wes Anderson's
movies to the Coen Brothers.

First thing is to register for an authentication key. I did that, and will store it
in a variable to make it easier to reuse.

```{r}
m_key <- "&api-key=24cf73bd0d79a879f2c726186680b1c7:4:73355908"
```

The next step is to build my request string in the form of a URI. Looking at the docs,
I can see it version should always be v2. I know I want json, so that goes after the search
part of the string. The question mark starts my query string, where I'll offer my search
terms.

```{r}
cb_uri <- "http://api.nytimes.com/svc/movies/v2/reviews/search.json?query=joel+ethan+coen"
```

Next, I need to combine that with my authentication key, and send the request.

```{r}
coen_bros <- fromJSON(paste0(cb_uri, m_key))
```

The response object is a list of 4 items. The one I'm interested in is the results data frame,
so I'll pull that out.

```{r}
coen_bros <- coen_bros$results
```

Now I just need to build another URI for the Wes Anderson Movies.

```{r}
wa_uri <- "http://api.nytimes.com/svc/movies/v2/reviews/search.json?query=wes+anderson"

wes <- fromJSON(paste0(wa_uri, m_key))
wes <- wes$results
```


If I look at the titles, I'll see that this search returns more than just movies that are by Wes Anderson.

```{r}
wes$display_title
```

So I'll isolate only the ones I know he directed.

```{r}
wes_df <- wes[1:8,]
```

Same thing for the Coen Brothers.

```{r}
coen_bros$display_title
coen_df <- coen_bros[1:17,]
```


There are more columns than I need, so I'll build new data frames that are a little leaner.

```{r}
wes_df <- data.frame(title = wes_df$display_title, 
                      rating = wes_df$mpaa_rating, 
                      critics_pick = wes_df$critics_pick, 
                      top1000 = wes_df$thousand_best)

coen_df <- data.frame(title = coen_df$display_title, 
                       rating = coen_df$mpaa_rating, 
                       critics_pick = coen_df$critics_pick, 
                       top1000 = coen_df$thousand_best)
```

Now I have clean, well formatted dataframes representing the NY Times opinion of each director.


The NY Times doesn't do scores, which while maybe artistically appropriate, is inconvenient for data analysis. Really the only thing I have to work with are the critics_pick and top1000 variables, which are both binary.

Comparing how many films of each were critics_pick yields this:

```{r}
table(coen_df$critics_pick)
table(wes_df$critics_pick)
```

This shows the NY Times like the Coen Brothers movies more often than they do Wes Anderson's. I
would agree with them on this.

