---
title: "IS607 Week 4 Assignment"
author: "Chris Fenton"
date: "September 18, 2015"
output: html_document
---

These problems come from the book **Automated Data Collection with R** 1st edition, Chapter 8 problems 4-8 on pages 217-218.

Before working, I'll load the "stringr" package to use with the regular
expressions.

```{r}
library(stringr)
```

###4. 

*Describe the types of strings that conform to the following regular
expressions and construct an example that is matched by the regular
expression.*

(a) [0-9]+\\\\$

Answer: this expression matches a digit sequence of any length followed
by a literal dollar sign, since the double backslash escapes the metacharacter
meaning typically associated with the dollar sign (which is end of sentence).

The string "398\$hello" returns "398$":
```{R}
str_extract("398$hello", "[0-9]+\\$")
```

The string "398" returns "N/A", showing again that the $ is literal:
```{R}
str_extract("398", "[0-9]+\\$")
```

(b) \\\\b[a-z]{1,4}\\\\b

Answer: The expression is bookended by word edge characters, so it is looking
for a single word. The word consists of any lowercase letter, and has to be
between 1 and 4 characters long.


```{r}
str_extract("I am a real american", "\\b[a-z]{1,4}\\b")

unlist(str_extract_all("I am a real american", "\\b[a-z]{1,4}\\b"))
```

(c) .*?\\\\.txt$

This expression looks for any series of characters followed by ".txt" at the end of the string. The question mark is necessary to make the expression "ungreedy", thus
stopping at the next match of the expression.

```{r}
str_extract("txt", ".*?\\.txt$")
str_extract(".txt", ".*?\\.txt$")
str_extract("jabroni.txt", ".*?\\.txt$")
str_extract("know your role blvd jabroni.txt", ".*?\\.txt$")
```

(d) \\\\d{2}/\\\\d{2}/\\\\d{4}

This looks like a date format, almost. It's the same form, however the digits
between the forward slashes are not restricted to valid date ranges.

```{r}
str_extract("12/12/1999", "\\d{2}/\\d{2}/\\d{4}")
str_extract("30/30/1999", "\\d{2}/\\d{2}/\\d{4}")
str_extract("dd/dd/dddd", "\\d{2}/\\d{2}/\\d{4}")
```

(e) <(.+?)>.+?</\\\\1>

This regex matches to a html/xml style tag. The first tag is any sequence of
characters enclosed in < > braces, and the second tag are those same characters
with a leading forward slash. There must be some character in between.

```{r}
str_extract("<head> </head>", "<(.+?)>.+?</\\1>")
str_extract("<title>ol' chunk of coal</title>", "<(.+?)>.+?</\\1>")
str_extract("<head>won't work<head>", "<(.+?)>.+?</\\1>")
str_extract("<head>this neither</title>", "<(.+?)>.+?</\\1>")
```

###5.

*Rewrite the expression [0-9]+\\\\$ in a way that all elements are altered
but the expression performs the same task.*

Answer: \\\\d{1,}[$]

```{r}
str_extract("398$hello", "\\d{1,}[$]")
str_extract("398", "\\d{1,}[$]")
```

###6.

*Consider the mail address chunkylover53[at]aol[dot]com*

(a) *Transform the string to a standard mail format using regular expressions*

```{r}
email <- "chunkylover53[at]aol[dot]com"
email <- str_replace(email, "\\[at\\]", "@" )
email <- str_replace(email, "\\[dot\\]", ".")
email
```

(b) *Imagine we are trying to extract the digits in the mail address. To do so
we write the expression [:digit:]. Explain why this fails and correct the 
expression.*

Answer: This fails because predefined character classes must be enclosed in
brackets. If they aren't, R will make a character class of ":digit".

```{r}
str_extract_all(email, "[[:digit:]]")
```

(c) *Instead of using the predefined character classes, we would like to use
the predefined symbols to extract the digits in the mail address. To do so we
write the expression \\\\D. Explain why this fails and correct the expression.*

Answer: \\\\D looks at everything **but** digits. \\\\d returns digits.

```{r}
str_extract_all(email, "\\d")
```

###7. 

*Consider the string <title>+++BREAKING NEWS+++</title>. We would would like
to extract the first HTML tag. To do so we write the regular expression <.+>..
Explain why this fails and correct the expression.*

Answer: This string fails because it watched the movie *Wall Street* one too
many times: it's too greedy. Adding the ? character after the  plus sign
will make take the shortest possible string that will satisfy the conditions.

```{r}
html_string <- "<title>+++BREAKING NEWS+++</title>"
str_extract(html_string, "<.+>")
str_extract(html_string, "<.+?>")
```

###8.

*Consider the string "(5-3)^2 = 5^2-2\*5\*3+ 3^2 conforms to the binomial 
theorem" We would like to extract the formula in the string. To do so we 
write the regular expression [^0-9=+*()]+. Explain why this fails and
correct the expression.*

Answer: Putting the character "^" at the beginning inverses the rest of the character class's contents. "-" is not really a member of the character class, since it retains its metacharacter status. If we move the carrot and the 
dash to the end, that makes it so the class is no longer inverted and
the dash is included in the character class.

```{r}
binomial_str <- "(5-3)^2=5^2-2*5*3+3^2 conforms to the binomial theorem."
str_extract(binomial_str, "[^0-9=+*()]+")
str_extract(binomial_str, "[0-9=+*()^-]+")

```