---
title: "IS 607 Week 6 Assignment"
author: "Chris Fenton"
date: "October 4, 2015"
output: html_document
---

The goal here is to transform a wide data set into a "tidy" one. A tidy data
set is defined as one where each variable is a column, each observation is a
row, and each type of observational unit is a table.

I'll be using the tidyr and dplyr packages.

```{r}
library(tidyr)
library(dplyr)
```

First I'll read in the data.

```{r}
w6 <- read.csv('week_6.csv')
```

get rid of the blank row and add in the other two missing variables.

The way the data is structured I can see that the "Alaska" and "AM West" 
observations span two lines. I'll add them to the 2nd line of each and 
get rid of the blank middle line.

```{r}
w6[2,1] <- 'ALASKA'
w6[5,1] <- 'AM WEST'
w6 <- filter(w6, w6[,1] != "")
```

Since there were no headers on iehter the airline or on time/delayed columns,
they were named X and X.1 by default. I will rename those something more 
usable.

```{r}
names(w6)[1] <- "Airline"
names(w6)[2] <- "timeliness"
```

If I look at the data...

```{r}
w6
```

...everything is filled in, so now I'm ready to transform via tidyr.


I'll use piping syntax because it makes reduces the amount of typing.
```{r}
w6_tidy <- w6 %>% gather(city, frequency, -Airline, -timeliness) %>%
                  spread(timeliness, frequency)
```

First I used gather() to take all columns other than airline and timeliness,
stack them in a new column called "city", and have the new value column called
frequency.

Next, I thought it would make sense to replace the frequency column with a
'delayed' column and a 'on time' column.

Here's the result:

```{r}
w6_tidy
```

I thought about combining the airline with the ontime and delayed columns, so
there would be a "Alaska On Time" column, "Alaska Delayed", and so forth. But
this seemed to combine two variables in a column, thus violating the definition
of 'tidy' data. 

However it seems to me that while there are good general rules
for what constitutes 'tidy' data, there are some areas where it is subjective,
and one is probably best served by thinking about making it 'tidy' in terms
of the anlysis they are trying to perform.

So with that in mind, I'll add some columns to measure relative delays
per city per airline.

```{r}
w6_tidy$total <- w6_tidy$`on time` + w6_tidy$delayed
w6_tidy$delay_pct <- w6_tidy$delayed / w6_tidy$total
```

Now I can compare how the airlines performed in any city I want to look at
directly.

```{r}
filter(w6_tidy, city == 'Los.Angeles')
```

So Alaska had relatively (and absolutely) fewer delays in Los Angeles.

If I use the arrange() function to sort by the city, I can compare all the
airlines at once:

```{r}
arrange(w6_tidy, city)
```

And it appears Alaska has relatively fewer delays in *every* city.
