---
title: "IS607 Project 1 - Chess"
author: "Chris Fenton"
date: "September 19, 2015"
output: html_document
---

###Introduction

The goal of this project is to take a text file that contains the results from
a chess tournament, parse through the file, extract relevant data, add
structure, and produce a .csv file.

The .csv file will contain the following variables:  
  * Player's Name  
  * Player's State  
  * Total Number of Points  
  * Player's Pre-Rating  
  * Average Pre Chess Rating of Opponents  
  
I will be using the 'stringr' package for regex, rather than working with the
default package.

```{r}
library(stringr)
```

###Preparing the Text

My first step is to load the file using the readLines function.

```{r warning=FALSE}
chess_table <- readLines("tournamentinfo.txt")
```

Looking at the text directly, there appears to be a good amount of structure
before I've done anything. Also, all the relevant information appears to be 
on one of two lines associated with each participant.

With that in mind, my strategy will be to create a data frame with one
row per participant, and add the two lines of raw text in strings as
additional columns for that participant. That way, it will be easy to associate
text with the participants.

I do this by applying the str_detect function from the stringr package. 
str_detect returns a logical vector that lets me identify whether or not
each line fits my criteria (defined via regex) of whether its a first line
or second line. I use lapply to try to keep things functional.

As for the expressions, these are fairly easy to write as the text had a
disnct pattern. The first type of line that contains data always begins with
a number of one or two digits, and is the only type of line that has this
pattern. The second type of line always begins with a pair of upper case
letters, and again is the only type of line that follows this pattern.

```{r}
first_line <- unlist(lapply(chess_table, str_detect, 
                              pattern = "^\\s+\\d{1,2}.+"))
second_line <- unlist(lapply(chess_table, str_detect,
                              pattern = "^\\s+[[:upper:]]{2}.+"))

chess_data <- data.frame(chess_table, 
                         first_line, 
                         second_line, 
                         stringsAsFactors = FALSE)
```

Now that I have a dataframe that identifies which pattern each line in the
text file follows, I will build another dataframe that actually contains the
data, along with 2 patterns of line for eacth participant.

```{r}
first_vector <- subset(chess_data$chess_table, first_line == TRUE)
second_vector <- subset(chess_data$chess_table, second_line == TRUE)

tourny_data <- data.frame(ID = 1:64, 
                          line1 = first_vector, 
                          line2 = second_vector, 
                          stringsAsFactors = FALSE)
```

###Extracting the Data

With my dataframe (tourny_data) in hand, my plan of attack will be to identify a 
data element from the list in the introduction, examine the file to detect a 
pattern surrounding that data in the text, extract it using str_extract, and 
add a vector to tourny_data.

####Player Name

The players name consists of several separate
words of uppercase letters and hyphens. I can look for any number of words 
consisting of those characters followed by a space, capped by a word that
stops before any space (so there won't be a trailing space at the end). The
name is always on line 1.
```{r}
tourny_data[,"player_name"] <- str_extract(tourny_data$line1, 
                           "(\\b[[:upper:]-]+\\b\\s)+(\\b[[:upper:]-]+\\b){1}")
```

####Player State

The state is always the first two uppercase letters on line2.
```{r}
tourny_data[,"state"] <- str_extract(tourny_data$line2, "[[:upper:]]{2}" )
```

####Total Points

The total points are the only decimal numbers in the file, making it very easy
to extract. I also coerce the vector to "numeric"" before adding it to the 
dataframe, in case I want to measure it at all.
```{r}
tourny_data[,"points"] <- as.numeric(str_extract(tourny_data$line1, 
                                                  "\\d(.)\\d"))
```

####Pre-Rating

Finally, a challenge! (just kidding) Anyway, the pre-rating does present an issue
in that the pre-ratings are not formatted consistently. Most players have ratings
that are 3 or 4 digits and stand alone on line 2, but several have scores that 
are 3 or 4 digits, but are followed by a "P" and some more digits.

I will first handle the majority format. I use an element of regex here called
"lookbehind", which basically lets you detect a character without actually
matching it. This is very useful if you know the characters you are looking for
have a consistent border, which in my case are a blank space. Also, in
order to use this feature of regex in the stringr package, you have to wrap
the expression argument in a "perl()" function call, as this changes the regex
engine to perl.

For the rest of the expression, I'm just looking for 3 or 4 digits that end a word.
Again, I'll coerce to numeric for computational ease down the road.

```{r}
tourny_data[,"pre_rating"] <- as.numeric(str_extract(tourny_data$line2,
                                                     perl("(?<=\\s)(\\d){3,4}\\b")))
```

So now I've added a vector onto tourny_data that either contains the player's
pretournament rating, or NA. The NAs are actually useful because now I can
subset on them and handle the differing formats directly.

As mentioned before, the main distinguishing characteristic of these records
is that the score is followed by an uppercase "P". In a similar manner to
"lookbehind" above, I can use "lookahead" here for a "P" without actually
extracting it.

```{r}
tourny_data[is.na(tourny_data$pre_rating), 
                            "pre_rating"] <- as.numeric(
                                              str_extract(
                                                subset(
                                                  tourny_data$line2, 
                                                  is.na(tourny_data$pre_rating)), 
                                                perl("\\b\\d{3,4}(?=P)")))
```

####Average Opponent Score

This part of the problem is unique in that I need to do more than just find 
data within the text. I have to extract data from the text in order to 
reference other data, and than do a calculation.

First thing is to get the reference data, which are the opponent IDs. This
time I will use str_extract_all as I need more than one piece of data. Luckily,
the text is in a simple, consistent, unique pattern: any two digits followed
by a "|" on line 1. 

```{r}
opponents <- str_extract_all(tourny_data$line1, perl("(\\d){1,2}(?=[|])"))
opponents <- lapply(opponents, as.numeric)
```

So now I have an 'opponents' list, but it does not actually contain what I
need, it only tells me where to look for what I need. Since each vector in the
list will need to follow the same process, I find the best way forward is to 
write a function.

My function, called avg_score_lookup, takes a numeric vector as input.
It first measures the length of each vector, thus tallying the total games 
played by a player. Next, it instatiates an empty vector to keep track of the 
total pre tournament ratings of all opponents. 

A for loop runs through the input opponents vector, subsets on the tourney_
data using the opponents ID and adds the pre rating to the toal. Finally, the 
total is divided by the number of opponents and returned, providing the average 
opponent score.
```{r}
avg_score_lookup <- function(opponents) {
  num_of_opponents <- length(opponents)
  total_rating <- 0
  for (i in opponents){
      total_rating <- total_rating + tourny_data[i, "pre_rating"]
  }
  return(total_rating/num_of_opponents)
}
```

With this function in hand, I can apply it to my list of opponents and 
glue the resulting vector on to my data frame. 
```{r}
tourny_data[,"avg_opp_prerate"] <- unlist(lapply(opponents, avg_score_lookup))
```
###Finishing Up

So now I have all the required data in a single data frame. Before I write
the .csv file, I'll create a sperate data frame in order to remove the raw 
text and the ID numbers I didn't use.

```{r}
chess_csv <- tourny_data[,4:8]
```

I'll have a quick look at the head as a final check.

```{r}
head(chess_csv)
```

Looks good, so we're ready to write.

```{r}
write.csv(chess_csv, "chess_data.csv")
```

And that's the file.

###Some Analysis

It'd be a shame to do all that dirty work and not at least do a little bit of
explatory analysis.

First off, I wonder where the tournament was?

```{r}
table(chess_csv$state)
```

Looks like Detroit, perhaps?

Next, I'll look at the distribution of participant skill.

```{r}
summary(chess_csv$pre_rating)
hist(chess_csv$pre_rating, breaks = 50)
```

The distribution is multimodal, left skewed, and centered around 1400. According to some
quick online research, this is right in line with the USCF median, and a bit higher than
the USCF mean. 